Reducing RES Droughts through the integration of wind and solar PV

### 2025-04-01 Article/RES_droughts_Morin_ReEn_R2.tex
Changed R2 version to be the same as last Thrursday Evening

In addition to examining periods of low renewable electricity generation, several studies also explore the periods when the imbalance between renewable generation and electricity demand (residual load) is high. Raynaud et al.~\citep{raynaud2018drought} showed the difference between RES droughts and high residual load events in a hypothetical fully renewable system composed of wind, solar PV and run-of-the-river hydropower. Similarly, Allen and Otero~\citep{allen2023drought} also defined a standardised index based on meteorological droughts to address residual load, whose correlation to the electricity generation index is mostly negative (as expected, although quite low anticorrelations and even small positive correlations appear for some European countries). This index was also applied to the U.S. by Bracken et al.~\citep{bracken2024drought}, revealing a consistent increase in the RES drought magnitude when load is considered, despite showing differing results across regions.

In this paper, the focus is exclusively on renewable electricity generation, which allows us to maintain physical models that do not consider the behavioural influence of demand. A fixed threshold approach is used to define RES droughts, which facilitates consistent inter-comparison between scenarios with different installed wind and solar PV capacities.  The case study used in this paper is Ireland, a region where most RES generation comes from wind power and with ambitious targets for solar PV power expansion. This provides valuable insights into the potential benefits of adding solar PV installations in wind-dominated countries.

RES droughts are identified using onshore wind and solar PV CF time series. In this study, three different datasets are used and compared, all of which are driven by the ERA5 reanalysis~\citep{hersbach2020era5}. Two of the datasets are part of C3S Energy (C3SE), an energy-based operational dataset produced by the EU Copernicus Climate Change Service~\citep{dubus2023energy}. One of the C3SE datasets provides CF time series aggregated at the national scale, while the other provides the CF time series at each grid point, at the ERA5 resolution of 0.25°. The third dataset produced by the authors was generated using the Atlite model~\citep{hofman2021atlite}, which converts the ERA5 atmospheric data to a generation time series using specified wind turbine and PV panel models. Atlite is an open-source tool developed by PyPSA~\citep{hofman2021atlite} and has been used for estimating wind and solar PV generation in order to study RES droughts in Germany~\citep{mockert2023drought}.

Generic datasets for wind and solar PV CF are often used for the quantification of RES droughts. Despite undergoing a validation process, they are often not fully representative of each geographical location, and can show differences in the number of RES drought events~\citep{kies2021drought}. This study evaluates the skill of a dataset developed for the European region (C3SE) when applied to a specific country (Ireland). In particular, the analysis explores the impact of using a generic versus a tailored dataset on RES drought assessments, in the context of a transition from a wind-dominated system to one with a greater share of solar PV.

@@

The EU Copernicus Climate Change Service developed the C3S-Energy (C3SE) renewable energy dataset for Europe~\citep{dubus2023energy}, using ERA5 atmospheric variables and weather-to-energy models. This dataset provides hourly CF for wind and solar PV power from 1979 to the present. The data are available on the same grid as the ERA5 data, which has a horizontal resolution of 0.25°. The time series are also available for download at two aggregated scales: regional (NUTS 2) and national.

@@

This study analyses RES droughts across the island of Ireland using onshore wind and solar PV CF time series from three datasets: two from C3SE, based on national-level data (C3S NAT) and grid-level data (C3S GRD), and one derived from the Atlite model (ATL). Fig.~\ref{fig:violin_plots} presents the statistics of the three datasets in violin plots for wind and solar PV. These plots illustrate the density of CF values over time, highlighting differences between the datasets and their alignment with observed data.

\begin{figure}[ht!]
	\centering
	\includegraphics[width=\textwidth]{violin_plots_verification.pdf}
	\caption{Violin plots of CF distributions for wind a) and solar PV b) datasets. Each violin represents the distribution of CF over time for different datasets: Observations (black), ATL (red), C3S GRD (blue), and C3S NAT (purple). The mean value for each dataset is marked with a green horizontal line. The red line indicates the threshold of 0.1 used in the study to identify RES droughts}
	\label{fig:violin_plots}
\end{figure}

@@

This verification for wind generation data highlights the importance of selecting a representative wind turbine power curve for the region being analysed. The ATL dataset, which uses a representative wind turbine power curve, is skilled at reproducing wind CF and RES droughts over Ireland. On the other hand, the power curve used for both C3S GRD and C3S NAT is not representative for Ireland, as it severely overestimates generation, underestimating the occurrence of RES droughts. This highlights a problem with using generalised datasets for analysing RES droughts: biases severely affect their ability to accurately reproduce RES drought events. The skill scores for the three datasets (Tab.~\ref{tab:wind_skill_scores}) show only a small difference in their ability to reproduce the changes in CF, as seen by their similar CC scores. However, their ability to reproduce the actual CF values is much lower than that of ATL, with RMSE scores almost four times bigger for the two C3S datasets. There is a clear bias towards an overestimation of CF, seen in the MBE values, which leads to the underestimation of RES droughts. This highlights the need to use regionally verified models to assess RES droughts.

@@

The first part of the analysis examines the annual number of RES drought events. When only wind energy is considered (Fig.~\ref{fig:boxplot_number_events}a), the number of RES drought events decreases as the duration range increases, with very few events lasting more than seven days. In contrast, for solar PV energy (Fig.~\ref{fig:boxplot_number_events}b), RES drought frequency declines from one to eight days and then slightly increases for longer durations. This behaviour is attributable to Ireland's high-latitude location, where reduced sunlight in winter (from November to March) leads to consistently low solar PV output.

Moreover, the comparison between wind and solar PV results indicates that the median, first, and third quartiles for solar PV are consistently higher than or equal to those for wind. This is expected, given that solar PV generation is inherently lower, zero at night, and limited by the solar cycle. When wind and solar PV are combined under the 91W-9PV scenario (Fig.~\ref{fig:boxplot_number_events}c), the results closely mirror those of wind alone, due to the dominance of wind power in the current energy mix. However, in the 57W-43PV scenario (Fig.~\ref{fig:boxplot_number_events}d), a marked reduction in RES drought events is observed across all datasets, with a decrease of the total number of events of 56\% for ATL, 52\% for C3S GRD, and 50\% for C3S NAT, demonstrating the beneficial effects of a more balanced energy mix.

@@

This study has explored the characterisation of RES droughts in the transition from a wind-dominated system to a more balanced system with integrated solar PV, based on the real case of Ireland. Three different datasets were compared over a 45-year period: one created using a regionally validated model and two derived from a generic dataset for Europe, C3S-Energy. The two datasets derived from C3S-Energy present different approaches, with one using large-scale aggregated information only, and the other one including the locations of farms as well. The regionally validated model considered the locations of farms as well as tailored wind and solar PV models selected to best represent the actual generation in Ireland.

Our results show the limitations in the quantification of RES droughts present in datasets that have not undergone regional validation. The three datasets used in this study are able to capture overall trends in RES drought occurrence such as the seasonal cycle or the effect of increasing the share of solar PV. However, significant differences in the quantitative values, particularly the extremes, emerge when using non-validated datasets for the study of RES droughts. This finding highlights that using a non-validated dataset can lead to undersized reserve capacity, with the associated negative consequences for grid stability and security of supply.

This study has also revealed that differences in the wind turbine power curves and solar PV panel models have a stronger influence on the estimation of RES droughts than the consideration of RES farm locations. The two datasets derived from C3S-Energy consistently underestimated the number of wind drought events and the frequency of extremes when compared to the regionally validated dataset with a specifically selected wind turbine power curve. This suggests that a meticulous selection of the wind turbine power curve to match observed data is crucial for accurately quantifying RES drought risks, thereby supporting more effective energy system planning.

Finally, the effect of the integration of solar PV in a wind-dominated system on RES droughts has been explored in a real-case setting based on Ireland. Our analysis has demonstrated that transitioning to a system with similar amounts of wind and solar PV reduces the frequency, duration, and seasonal variability of RES drought events. This improvement is attributed to the complementary nature of wind and solar PV generation, as solar PV typically peaks in summer while wind generation is more consistent in winter. However, this integration is unable to counter the critical winter RES droughts, which coincide with the strongest electricity demand in Northern European countries like Ireland. Still, a more diversified renewable energy mix mitigates extreme RES drought conditions and enhances overall system resilience.

The results presented in this study have four main limitations. First, the presented study uses a fixed threshold to define RES drought events, but other methods could yield different results, even though the main takeaways would be expected to be the same. Second, the definition of RES droughts based on generation does not consider the mismatch between renewable generation and demand, which may be of interest to system operators, as these events put large amounts of strain on the system. Third, the availability of solar PV data is limited to a relatively short time period, as recent expansions in installed capacity have significantly changed the generation landscape. Lastly, the source for weather data is ERA5, which is among the best reanalysis datasets for renewable energy applications, but still comes in a limited spatial resolution.

>>>>

In addition to examining periods of low renewable electricity generation, several studies also explore the periods when the imbalance between renewable generation and electricity demand (residual load) is high. Raynaud et al.~\citep{raynaud2018drought} defined both energy production and energy supply droughts, and showed the difference in their patterns in a hypothetical fully renewable system composed of wind, PV and run-of-the-river hydropower. Similarly, Allen and Otero~\citep{allen2023drought} also defined a standardised index based on meteorological droughts to address residual load, whose correlation to the energy production index is mostly negative (as expected, although quite low anticorrelations and even small positive correlations appear for some European countries). This index was also applied to the U.S. by Bracken et al~\citep{bracken2024drought}, revealing a consistent increase in the drought magnitude when load is considered, despite showing differing results across regions.

In this paper, the focus is exclusively on renewable electricity generation, which allows us to maintain physical models that do not consider the behavioural influence of demand, whose role will be addressed in the discussion. A fixed threshold approach is used to define RES droughts, which facilitates consistent inter-comparison between scenarios with different installed wind and solar PV capacities. The case study used in this paper is Ireland, a region with a strong reliance on wind power and ambitious targets for solar PV power expansion. This provides valuable insights into the potential benefits of diversifying the renewable energy mix on RES droughts in the context of realistic scenarios.

RES droughts are identified using onshore wind and solar PV CF time series. In this study, three different datasets are used and compared, all of which are driven by the ERA5 reanalysis~\citep{hersbach2020era5}. Two of the datasets are part of C3S Energy (C3SE), an energy-based operational dataset produced by the EU Copernicus Climate Change Service~\citep{dubus2023energy}. One of the C3SE datasets provides CF time series aggregated at the national scale, while the other provides the CF time series at each grid point, at the ERA5 resolution of 0.25°. The third dataset was generated using the Atlite model~\citep{hofman2021atlite}, which converts the ERA5 atmospheric data to a generation time series using specified wind turbine and PV panel models. Atlite is an open-source tool developed by PyPSA~\citep{hofman2021atlite} and has been used for estimating wind and solar PV generation in order to study RES droughts~\citep{mockert2023drought}.

Generic datasets for wind and solar PV CF are often used for the quantification of RES droughts. Despite undergoing a validation process, they are often not fully representative of each geographical location, and can show differences in the number of RES drought events \citep{kies2021drought}. In this work, we quantify the skill of a dataset developed for the European region (C3SE), when used for a specific country (Ireland). In particular, we investigate the impact of a generic versus a tailored dataset on the analysis of RES droughts, in the context of a transition from a wind-dominated system to one with a larger share of solar PV.

@@

The EU Copernicus Climate Change Service developed the C3S-Energy (C3SE) renewable energy dataset for Europe~\citep{dubus2023energy}, using ERA5 atmospheric variables and weather-to-energy models. This dataset provides hourly CF for wind and solar PV energy from 1979 to the present. The data are available on the same grid as the ERA5 data, which has a horizontal resolution of 0.25°. The time series are also available for download at two aggregated scales: regional (NUTS 2) and national.

@@

This study uses onshore wind and solar PV CF time series from three datasets to analyse RES droughts across the island of Ireland. Data downloaded from C3SE were used to obtain two datasets: one based on national-level data (C3S~NAT), and another on grid-level data (C3S~GRD). The third dataset was computed using the Atlite model (ATL).

@@

This verification for wind generation data highlights the importance of selecting a representative wind turbine power curve for the region being analysed. The ATL dataset, which uses a representative wind turbine power curve, is skilled at reproducing wind CF and droughts over Ireland. On the other hand, the power curve used for both C3S GRID and C3S NAT is not representative for Ireland, as it severely overestimates generation, underestimating the occurrence of RES droughts. This highlights a problem with using generalised datasets for analysing RES droughts: biases severely affect their ability to accurately reproduce RES drought events. The skill scores for the three datasets (Tab.~\ref{tab:wind_skill_scores}) show only a small difference in their ability to reproduce the changes in CF, as seen by their similar CC scores. However, their ability to reproduce the actual CF values is much lower than that of ATL, with RMSE scores almost four times bigger for the two C3S datasets. There is a clear bias towards an overestimation of CF, seen in the MBE values, which leads to the underestimation of droughts. This highlights the need to use regionally verified models to assess RES droughts.

@@

The first part of the analysis examines the annual number of RES drought events. When only wind energy is considered (Fig.\ref{fig:boxplot_number_events}a), the number of drought events decreases as the duration range increases, with very few events lasting more than seven days. In contrast, for solar PV energy (Fig.\ref{fig:boxplot_number_events}b), drought frequency declines from one to eight days and then slightly increases for longer durations. This behaviour is attributable to Ireland's high-latitude location, where reduced sunlight in winter (from November to March) leads to consistently low solar PV output.

Moreover, the comparison between wind and solar PV results indicates that the median, first, and third quartiles for solar PV are consistently higher than or equal to those for wind. This is expected, given that solar PV generation is inherently lower, zero at night, and limited by the solar cycle. When wind and solar PV are combined under the 91W-9PV scenario (Fig.\ref{fig:boxplot_number_events}c), the results closely mirror those of wind alone, due to the dominance of wind power in the current energy mix. However, in the 57W-43PV scenario (Fig.\ref{fig:boxplot_number_events}d), a marked reduction in drought events is observed across all datasets, with a decrease of the total number of events of 56\% for ATL, 52\% for C3S GRD, and 50\% for C3S NAT, demonstrating the beneficial effects of a more balanced energy mix.

@@

This study has explored the characterisation of RES droughts in the real-life transition from a wind-dominated system to a more balanced system with integrated solar PV. This has been done through the comparison over a 45-year period of three different datasets: one based on a hand-crafted validated model and two based on a generic model, C3S-Energy. The generic model has two versions, one using large-scale aggregated information, and one which includes the locations of farms as well (also considered in the validated model).

Our results show that generic datasets present limitations in the quantification of RES droughts. Although the three datasets used capture overall trends in drought occurrence, significant differences emerge when using non-tailored data for the study of RES droughts. This finding highlights that the choice of dataset and its underlying assumptions can lead to consequential differences in estimated RES drought characteristics, emphasising the need for datasets that are specifically designed for extreme event analysis.

This study reveals that differences in model parametrisation, particularly in the representation of wind turbine power curves and solar PV panel characteristics, have a stronger influence on the drought estimates than the inclusion of RES farm locations. The use of a validated dataset with a carefully selected wind turbine power curve consistently produced higher return periods and a greater number of drought events for wind energy than the datasets derived from C3S-Energy. This suggests that fine-tuning model parameters to match observed data is crucial for accurately quantifying RES drought risks, thereby supporting more effective energy system planning.

Finally, the effect on RES droughts of the integration of solar PV in a wind-dominated system has been explored in a real-case setting. In the presented example of Ireland, the analysis has demonstrated that transitioning to a more balanced system with similar amounts of wind and solar PV markedly reduces the frequency, duration, and seasonal variability of drought events. This improvement is attributed to the complementary nature of wind and solar PV generation, as solar PV typically peaks in summer while wind generation is more consistent in winter. Thus, a more diversified renewable energy mix not only mitigates extreme drought conditions but also enhances overall system resilience, providing valuable insights for policymakers tasked with ensuring energy security.

The results presented in this study have several limitations. Although ERA5 is among the best reanalysis datasets for renewable energy analysis, it still presents some biases, and its resolution may not capture local-scale phenomena. This is especially limiting if individual farms are considered, but its still shows good skill at statistically representing nation-wide behaviours considering the distribution of farms. Moreover, the methodology employs a fixed threshold to define RES drought events, which is necessary for comparing the three datasets considering only weather-derived drivers, but does not account for demand variations. Consequently, while this approach enables a consistent inter-comparison, it will not signal events that are most critical for power system operations for their mismatch of RES generation and demand.
